Investigation of Limited Detection Schemes for Light Scattering of Optically Trapped Asymmetric Particles

### 2022-12-06 Manuscript_rev1.tex
Add files via upload

\usepackage{amssymb, amsthm}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage{babel}
\usepackage{wrapfig}
\usepackage{geometry}
\usepackage{fancyref}
\usepackage{lineno}
\usepackage{color}
\usepackage{nomencl}

@@

\nomenclature{$\bar{\bf s}$}{Dimer's instantaneous orientation}
\nomenclature{$\hat{\bf n}_\alpha$}{Reference Orientation $\alpha$}

%\linenumbers

@@

To demonstrate the analysis, we study a simple, illustrative example
of a trapped entity more complex than a single sphere (i.e.\ a trapped
asymmetric dimer).  We explore how to optimise the experimental
measurement to obtain the best estimate of the dimer's instantaneous
orientation, as a further paradigmatic example of interpretation of
the scattering data to obtain dynamical information on the trapped
entity.  There is a paucity of literature on measuring the orientation
of trapped non-spherical particles till the recent attempt in
Ref.~\citenum{raudsepp2022estimating} where imaging was employed to
study the orientation of trapped dimers and a resolution of () was
achieved.  Here, we simulate the Brownian motion of a dimer and
calculate the resulting scattered light, and show how Bayesian
inference can be used to optimise extraction of the true dimer
orientation from the light scattering signals given a choice of
scattering measurement angles.

@@

@@

We use the Brownian OT package developed by Fung~\textit{et~al}.
\cite{Vigilante2020Brownian_OT} to simulate the motion of a asymmetric
dimer within an optical trap.  Brownian OT combines MSTM
\cite{Mishchenko1996MSTM} and ``Optical Tweezer Toolbox''
(\textit{ott}) \cite{Lenton2020} to simulate the motion of arbitrary
shaped sphere clusters.  We simulated the motion of silica dimer
trapped in a $5$ mW Gaussian beam; the dimer was comprised of two
spheres with radii $1\,{\rm \mu m}$ and $ 0.5\,{\rm \mu m}$,
respectively.  We ran our simulation over the span of 10 seconds, with
Brownian OT calculating the dimer's orientation and position every
$1 \ ms$.

@@

Our goal is to determine the orientation of our trapper dimer based on the scattering measurements $I(\hat{n}, \ \theta_k)$ collected from three detection fibres situated at $\theta_1,\ \theta_2, \ \& \ \theta_3$. Rather than get an exact estimate of the dimer's orientation the possible orientation space is discretised into $\textit{n}_ref \ = \ 30$ possible reference orientations $\hat{\bf n}_{\alpha}$ that are evenly distributed on a unit sphere \cite{Rey2006}. Using MSTM we compute the raw intensities that would be collected from the dimer in each reference orientation $I(\hat{\bf n}_{\alpha}, \theta_k)$. 
%
These raw intensities are normalised according to
\begin{align}
\label{eq:scale}
  y_k(\hat{\bf n}_\alpha)
  = 
  \frac{I(\hat{\bf n}_\alpha, \theta_k) - \langle I(\hat{\bf n},\theta_k) \rangle } 
  {\langle I^2(\hat{\bf n},\theta_k) \rangle -\langle I(\hat{\bf n}, \theta_k)\rangle^2}
\end{align}
where the denominator is simply the standard deviation for our values
of $I(\hat{\bf n}, \theta_k)$.

The reference orientations, raw intensities, and scaled signals are given in Tables~\ref{tab:A1} and \ref{tab:A2}.
%
In practice, $y_k(\bar{\bf s})$ will have an inherent amount of
uncertainty in our measurement due to noise from our detector.  We can compare any instantaneous signal $y_1, y_2, y_3$ to our reference signals by assuming that the measurement uncertainty for each detection fibre is Gaussian and is centred about $y_k(\bar{\bf n})$.
%
Combining the signals from all three detection angles provides us with the conditional probability that, should the dimer be in orientation
$\hat{\bf n}_\alpha$, it would produce a signal ($y_1$, $y_2$, $y_3$):
\begin{align}
\label{eq:conditional}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  = \prod^3_{k=1}
(2\pi\sigma_k^2)^{-1/2} 
e^{-(y_{k}-y_{k}(\hat{\bf n}_\alpha))^2/2\sigma_k^2}
\end{align}
where $\sigma_k$ is related to the uncertainty in the detected signal
from our dimer $y_k(\bar{\bf s})$.

@@

Equation~\eqref{eq:conditional} gives the probability that a signal
($y_1$, $y_2$, $y_3$) will be measured if a dimer is in reference
orientation $\hat{\rm n}_\alpha$.
%
Using Bayes' theorem, the probability that the dimer is in reference
orientation $\hat{\bf n}_\alpha$, given that the signal ($y_1$, $y_2$,
$y_3$) is measured, can be estimated:
%.
\begin{align}
  \label{eq:bayes}
  p(\hat{\bf n}_\alpha| y_1,y_2,y_3)
  &=
    \frac{p(y_1,y_2,y_3|\hat{\bf n}_\alpha)
    p(\hat{\bf n}_\alpha)}{p(y_1, y_2, y_3)}
\end{align}
where $p(\hat{\bf n}_\alpha)$ and $p(y_1, y_2, y_3)$ are the prior
estimates of the distributions of particle orientations and
instantaneous signals, respectively.
%
Without any prior evidence we assume that the orientation priori of our dimer $p(\hat{\bf n}_{\alpha})$ is uniform. Knowledge of the dimer from previous measurement can be used to inform our estimate of $p(\hat{\bf n}_{\alpha})$, covered in Section~\ref{sec:test}.  The latter prior $p(y)$ is the probability of measuring a signal ($y_1$, $y_2$, $y_3$).  This is given by taking the discrete integral over collection of reference orientations.
\begin{align}
  p(y_1, y_2, y_3)
  =
  \sum_{\alpha=1}^{n_{\rm ref}}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  p(\hat{\bf n}_\alpha)
\end{align}
From \eqref{eq:bayes} we get a mass probability distribution that denotes the probability that our dimer is in orientation $\hat{\bf n}_{\alpha}$ given a signal ($y_1$, $y_2$, $y_3$). In order to evaluate the accuracy of our method, we tested it using a Brownian dynamics simulation of the dimer (see Section~\ref{sec:brownian}).

@@

To measure the error in the model's estimate the distribution from Eq.~\eqref{eq:bayes} and an ideal result, $p_ideal$, were compared.  An ideal result would be one that is 0 for every $\hat{\bf n}$ apart from $\hat{\bf n}_{best}$.
\begin{align}
	p_{ideal} = 
	\begin{cases}
		1 & \text{when $\hat{\bf n}_\alpha$ = $\hat{\bf n}_{best}$}\\
		0 & \text{anywhere else}
	\end{cases}
\end{align}
The distribution from Eq.~\eqref{eq:bayes} will assign some non-zero probability to every reference orientation. The model's error was measured by calculating the Kullback-Leibler divergence $K_l$ between the two distributions.
\begin{align}
	K_{l, \#}(p_{ideal}\parallel p(\hat{\bf n}_\alpha| y_1, y_2, y_3))
	= p_{ideal}\ln \left[\frac{p_{ideal}}{p(\hat{\bf n}| y_1,y_2,y_3))}
	\right]
	\label{eq;kullback}
\end{align}
where a larger value of $K_l$ indicates that our model is less confident in its prediction of the dimer's orientation. The divergence provides us a measure of how confident our measurement is. To better reflect the model's accuracy we implemented a penalty condition that worsens the divergence if our model estimates the wrong reference orientation:
\begin{align}
	p(\hat{\bf n}|y_1,y_2,y_3) = 
	\begin{cases}
		p(\hat{\bf n}_{estimate}|y_1,y_2,y_3) & 
		\text{when $\hat{\bf n}_{estimate}$ = $\hat{\bf n}_{best}$}\\
		p(\hat{\bf n}_{estimate}|y_1,y_2,y_3)-p(\hat{\bf n}_{best}|y_1,y_2,y_3) & \text{when $\hat{\bf n}_{estimate} \ne \hat{\bf n}_{best}$}
	\end{cases}
\end{align}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth]{./Images/fig3.png}
	\caption{Model's estimation of dimer orientation while assuming $p(\hat{\bf n}_\alpha)$ to be uniform.}
	\label{fig:uniform}
\end{figure}

The reasoning for this lies with the fact that differences in reference signals and reference orientation spacing are not directly correlated with each other. In some cases, two reference orientation will have identical signal responses. To remedy this we need to account for the fact that a trapped dimer's motion i limited due to the trap stiffness. Therefore, the prior of the reference orientations $p(\hat{\bf n}_\alpha)$ was defined at each time step by accounting for the physical distance between our previous estimate $\hat{\bf n}_{\alpha}(t - \Delta t)$ and each reference orientation $\hat{\bf n}_\alpha(t)$:

\begin{align}
  r_i(t, \hat{\bf n}_\alpha)
  &= (\hat{\bf n}_{\alpha}(t) \cdot \hat{\bf n}_\alpha(t-\Delta t))
  \\
  p_i(\hat{\bf n})
  &= \frac{e^{\beta (r_i(t,\hat{\bf n}_i)-1)}}
{\sum_{j=1}^{n_{\rm ref}}e^{\beta(r_j(t, \hat{\bf n}_\alpha)-1)}}
\end{align}

>>>>

\usepackage{amssymb, amsthm}
\usepackage{amsmath}
\usepackage{subcaption}
\usepackage[T1]{fontenc}
\usepackage{babel}
\usepackage{wrapfig}
\usepackage{geometry}
\usepackage{fancyref}
\usepackage{lineno}
\usepackage{color}
\usepackage{nomencl}
\makenomenclature

@@

\linenumbers

@@

To demonstrate the analysis, we study a simple, illustrative example
of a trapped entity more complex than a single sphere (i.e.\ a trapped
asymmetric dimer).  We explore how to optimise the experimental
measurement to obtain the best estimate of the dimer's instantaneous
orientation, as a further paradigmatic example of interpretation of
the scattering data to obtain dynamical information on the trapped
entity.  There is a paucity of literature on measuring the orientation
of trapped non-spherical particles till the recent attempt in
Ref.~\cite{raudsepp2022estimating} where imaging was employed to
study the orientation of trapped dimers and a resolution of () was
achieved.  Here, we simulate the Brownian motion of a dimer and
calculate the resulting scattered light, and show how Bayesian
inference can be used to optimise extraction of the true dimer
orientation from the light scattering signals given a choice of
scattering measurement angles.

@@

\printnomenclature

@@

We use the Brownian OT package developed by Fung~\textit{et~al}. \cite{Vigilante2020Brownian_OT} to simulate the motion of a asymmetric dimer within an optical trap.  Brownian OT combines MSTM \cite{Mishchenko1996MSTM} and ``Optical Tweezer Toolbox'' (\textit{ott}) \cite{Lenton2020} to simulate the motion of arbitrary
shaped sphere clusters. We simulated the motion of a silica dimer
trapped in a $5$ mW Gaussian beam; the dimer was comprised of two tangent spheres with radii $1\,{\rm \mu m}$ and $ 0.5\,{\rm \mu m}$, respectively.  We ran our simulation over the span of 10 seconds, with Brownian OT calculating the dimer's orientation and position every $1 \ ms$.

@@

Our goal is to determine the orientation of our trapper dimer based on the scattering measurements $I(\hat{n}, \ \theta_k)$ collected from three detection fibres situated at $\theta_1,\ \theta_2, \ \& \ \theta_3$. Rather than get an exact estimate of the dimer's orientation the possible orientation space is discretised into $\textit{n}_{ref} \ = \ 30$ possible reference orientations $\hat{\bf n}_{\alpha}$ that are evenly distributed on a unit sphere \cite{Rey2006}. Using MSTM we compute the raw intensities that would be collected from the dimer in each reference orientation $I(\hat{\bf n}_{\alpha}, \theta_k)$. 
%
These raw intensities are normalised according to
\begin{align}
\label{eq:scale}
  y_k(\hat{\bf n}_\alpha)
  = 
  \frac{I(\hat{\bf n}_\alpha, \theta_k) - \langle I(\hat{\bf n},\theta_k) \rangle } 
  {\langle I^2(\hat{\bf n},\theta_k) \rangle -\langle I(\hat{\bf n}, \theta_k)\rangle^2}
\end{align}
where the denominator is simply the standard deviation for our values of $I(\hat{\bf n}, \theta_k)$. The reference orientations, raw intensities, and scaled signals are given in Tables~\ref{tab:A1} and \ref{tab:A2}.
%
In practice, $y_k(\bar{\bf s})$ will have an inherent amount of uncertainty in our measurement due to noise from our detector.  We can compare any instantaneous signal $y_1, y_2, y_3$ to our reference signals by assuming that the measurement uncertainty for each detection fibre is Gaussian and is centred about $y_k(\bar{\bf n})$.
%
Combining the signals from all three detection angles provides us with the conditional probability that, should the dimer be in orientation
$\hat{\bf n}_\alpha$, it would produce a signal ($y_1$, $y_2$, $y_3$):
\begin{align}
\label{eq:conditional}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  = \prod^3_{k=1}
(2\pi\sigma_k^2)^{-1/2} 
e^{-(y_{k}-y_{k}(\hat{\bf n}_\alpha))^2/2\sigma_k^2}
\end{align}
where $\sigma_k$ is related to the uncertainty in the detected signal
from our dimer $y_k(\bar{\bf s})$.

@@

Equation~\eqref{eq:conditional} gives the probability that a signal
($y_1$, $y_2$, $y_3$) will be measured if a dimer is in reference
orientation $\hat{\rm n}_\alpha$.
Using Bayes' theorem, the probability that the dimer is in reference orientation $\hat{\bf n}_\alpha$, given that the signal ($y_1$, $y_2$,
$y_3$) is measured, can be estimated:
%.
\begin{align}
  \label{eq:bayes}
  p(\hat{\bf n}_\alpha| y_1,y_2,y_3)
  &=
    \frac{p(y_1,y_2,y_3|\hat{\bf n}_\alpha)
    p(\hat{\bf n}_\alpha)}{p(y_1, y_2, y_3)}
\end{align}
where $p(\hat{\bf n}_\alpha)$ and $p(y_1, y_2, y_3)$ are the prior
estimates of the distributions of particle orientations and
instantaneous signals, respectively.
%
Without any prior evidence we assume that the orientation priori of our dimer
$p(\hat{\bf n}_{\alpha})$ is uniform. Knowledge of the dimer from previous
measurement can be used to inform our estimate of $p(\hat{\bf n}_{\alpha})$, covered in Section~\ref{sec:test}.  The latter prior $p(y)$ is the probability of measuring a signal ($y_1$, $y_2$, $y_3$).  This is given by taking the discrete integral over collection of reference orientations.
\begin{align}
  p(y_1, y_2, y_3)
  =
  \sum_{\alpha=1}^{n_{\rm ref}}
  p(y_1, y_2, y_3|\hat{\bf n}_\alpha)
  p(\hat{\bf n}_\alpha)
\end{align}
From \eqref{eq:bayes} we get a mass probability distribution that denotes the
probability that our dimer is in orientation $\hat{\bf n}_{\alpha}$ given a
signal ($y_1$, $y_2$, $y_3$). In order to evaluate the accuracy of our method, we tested it using a Brownian dynamics simulation of the dimer (see Section~\ref{sec:brownian}).

@@

To measure the error in the model's estimate the distribution from
Eq.~\eqref{eq:bayes} and an ideal result, $p_ideal$, were compared.  An ideal result would be one that is 0 for every $\hat{\bf n}$ apart from $\hat{\bf n}_{best}$.
\begin{align}
	p_{best} = 
	\begin{cases}
		1 & \text{when $\hat{\bf n}_\alpha$ = $\hat{\bf n}_{best}$}\\
		0 & \text{anywhere else}
	\end{cases}
\end{align}
The distribution from Eq.~\eqref{eq:bayes} will assign some non-zero
probability to every reference orientation. The model's error was measured by calculating the Kullback-Leibler divergence $K_l$ between the two distributions.
\begin{align}
	K_{l, \#}(p_{best}\parallel p(\hat{\bf n}_\alpha| y_1, y_2, y_3))
	= p_{best}\ln \left[\frac{p_{best}}{p(\hat{\bf n}| y_1,y_2,y_3))}
	\right]
	\label{eq;kullback}
\end{align}
where a larger value of $K_l$ indicates that our model is less confident in its 
prediction of the dimer's orientation. The divergence provides us a measure of how confident our measurement is. To better reflect the model's accuracy we implemented a penalty condition that worsens the divergence if our model estimates the wrong reference orientation:
\begin{align}
	p(\hat{\bf n}|y_1,y_2,y_3) = 
	\begin{cases}
		p(\hat{\bf n}_{estimate}|y_1,y_2,y_3) & 
		\text{when $\hat{\bf n}_{estimate}$ = $\hat{\bf n}_{best}$}\\
		p(\hat{\bf n}_{estimate}|y_1,y_2,y_3)-p(\hat{\bf n}_{best}|y_1,y_2,y_3) & \text{when $\hat{\bf n}_{estimate} \ne \hat{\bf n}_{best}$}
	\end{cases}
\end{align}

@@

\begin{figure}[h]
	\centering
	\includegraphics[width=0.5\textwidth]{./Images/fig3.png}
	\caption{Model's estimation of dimer orientation while assuming $p(\hat{\bf n}_\alpha)$ to be uniform. Dotted line denotes maximum spacing between two neighbouring $\hat{\bf n}_\alpha$, see text for further description.}
	\label{fig:uniform}
\end{figure}

The reasoning for this lies with the fact that differences in reference signals and reference orientation spacing are not directly correlated with each other. In some cases, two reference orientation will have identical signal responses. To remedy this we need to account for the fact that a trapped dimer's motion i limited due to the trap stiffness. Therefore, the prior of the reference orientations $p(\hat{\bf n}_\alpha)$ was defined at each time step by accounting for the physical distance between our previous estimate $\hat{\bf n}_{est}(t - \Delta t)$ and each reference orientation $\hat{\bf n}_\alpha(t)$:

\begin{align}
  p(\hat{\bf n}_\alpha)
  &=\frac{e^{\beta (\hat{\bf n}_\alpha (t) 
  	\cdot \hat{\bf n}_{est}(t-\Delta t))}}
  {\sum_{j=1}^{n_{\rm ref}}
	e^{\beta (\hat{\bf n}_\alpha (t) 
	\cdot \hat{\bf n}_{est}(t-\Delta t)}}
\end{align}
